Exploring the evolution of stellar rotation using Galactic kinematics

### 2020-04-20 paper/ms63.tex
adding a section for activity lifetime

\racomment{
In order to measure a rotation period at all, there must be some magnetically
active regions (either bright plages or dark spots), of a reasonable size on
the surface of a star.
In other words, some minimum magnetic activity level is presumably required to
produce coherent light curve-variability, from which a rotation period can be
confidently measured.
In this case, the relatively sharp upper-edge of the rotation period
distribution seen in figure \ref{fig:vplot} may be caused by a finite active
lifetime of stars, \ie\ stars older than a critical age are no longer active
enough to produce periodic, high-amplitude variability.
The data presented in figure \ref{fig:vplot} indicate that this active
lifetime is a function of stellar mass, because the velocity dispersions of
stars along the upper edge of the rotation period distribution increase with
decreasing mass.

The mass dependence of magnetic activity lifetimes has been demonstrated
previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a {\it
mass-dependent} activity lifetime.
Figure \ref{fig:vplot} shows that the populations of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that \racomment{most of} the oldest stars with detectable
rotation periods are cooler than 4500 K, \ie\ these-low mass stars stay active
longer than more massive stars.
}
% a decrease in magnetic activity after stars exceed a certain age.
% If so, it is possible that the ages of stars near this upper-edge represent
% the activity lifetimes of stars.}
% % We therefore assume that stars {\it with} measured rotation periods are,
% on % average, more magnetically active than stars {\it without} measured
% rotation % periods.
% \citet{matt2015} found that a 4 Gyr gyrochrone, calculated with their model,
% coincided with the upper-envelope at 4 Gyrs and concluded that the
% upper-envelope of the \kepler\ rotation period distribution corresponds to a
% single-age population of around 4 Gyr.
% However, the velocity dispersion data presented in figure \ref{fig:vplot}
% suggest that the activity lifetime of stars is mass-dependent.
% In this case, stars near the upper-dge of the period distribution are not all
% the same age that later-type stars are in fact older than earlier types.
% }

\racomment{
    To investigate this idea further, we compared the velocity dispersions of
stars with {\it measured rotation periods} to the velocity dispersions of the
entire \kepler\ sample.
% We assumed that stars with measured rotation periods are more magnetically
% active than stars without measured rotation periods.
If stars with measured rotation periods are more magnetically active, and
younger, than stars without rotation period measurements, they should also
have smaller velocity dispersions.
% This means that stars with measured rotation periods should be kinematically
% younger on average than the entire \kepler\ sample.
% they are young, because G stars are only magnetically active when they are
% young, G stars with {\it measured periods} should have smaller velocity
% dispersions than {\it all} G stars.
To test this theory, we calculated the velocity dispersions for all stars in
the \kepler\ field (after removing visual binaries, subgiants, stars fainter
than 16th magnitude, and high Galactic latitude stars, following the method
described in section \ref{sec:method}).
We then compared these velocity dispersions, as a function of
\teff, to the velocity dispersions of stars with {\it measured} rotation
periods \citep[\ie\ stars that appear in table 1 of][]{mcquillan2014}.
We show the ratio of total \sigmavb\ to McQuillan-\sigmavb\ as a
function of \teff\ in figure \ref{fig:compare}.}
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that the oldest K and M dwarfs are represented in the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods.
In other words, G stars become magnetically inactive and have fewer active
surface regions {\it at a younger age than M dwarfs}.

>>>>

\racomment{

\subsection{The mass-dependence of magnetic activity lifetimes}
\label{sec:activity}

The mass dependence of magnetic activity lifetimes has been demonstrated
previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
kiman2019} and, if the detectability of a rotation period is considered to be
a magnetic activity proxy, then our results provide further evidence for a
mass-dependent activity lifetime.

In order to measure a rotation period at all, there must be some magnetically
active regions (either bright plages or dark spots), of a reasonable size on
the surface of a star.
In other words, some minimum magnetic activity level is presumably required to
produce coherent light curve-variability, from which a rotation period can be
confidently measured.
The relatively sharp upper-edge of the rotation period distribution seen in
figure \ref{fig:vplot} may therefore be caused by a finite active lifetime of
stars, \ie\ stars older than a critical age are no longer active enough to
produce periodic, high-amplitude variability.
% The data presented in figure \ref{fig:vplot} indicate that this active
% lifetime is a function of stellar mass, because the velocity dispersions of
% stars along the upper edge of the rotation period distribution increase with
% decreasing mass.
In figure \ref{fig:vplot}, the velocity dispersions of stars along the upper
edge of the rotation period distribution increase with decreasing mass,
indicating that magnetic activity lifetimes are mass-dependent.
Figure \ref{fig:vplot} shows that the populations of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that most of the oldest stars with detectable rotation periods
are cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
}
% a decrease in magnetic activity after stars exceed a certain age.
% If so, it is possible that the ages of stars near this upper-edge represent
% the activity lifetimes of stars.}
% % We therefore assume that stars {\it with} measured rotation periods are,
% on % average, more magnetically active than stars {\it without} measured
% rotation % periods.
% \citet{matt2015} found that a 4 Gyr gyrochrone, calculated with their model,
% coincided with the upper-envelope at 4 Gyrs and concluded that the
% upper-envelope of the \kepler\ rotation period distribution corresponds to a
% single-age population of around 4 Gyr.
% However, the velocity dispersion data presented in figure \ref{fig:vplot}
% suggest that the activity lifetime of stars is mass-dependent.
% In this case, stars near the upper-dge of the period distribution are not all
% the same age that later-type stars are in fact older than earlier types.
% }

\racomment{
    To investigate this idea further, we compared the velocity dispersions of
stars with {\it measured rotation periods} to the velocity dispersions of the
entire \kepler\ sample.
% We assumed that stars with measured rotation periods are more magnetically
% active than stars without measured rotation periods.
If stars with measured rotation periods are more magnetically active, and
younger, than stars without rotation period measurements, they should also
have smaller velocity dispersions.
% This means that stars with measured rotation periods should be kinematically
% younger on average than the entire \kepler\ sample.
% they are young, because G stars are only magnetically active when they are
% young, G stars with {\it measured periods} should have smaller velocity
% dispersions than {\it all} G stars.
To test this theory, we calculated the velocity dispersions for all stars in
the \kepler\ field (after removing visual binaries, subgiants, stars fainter
than 16th magnitude, and high Galactic latitude stars, following the method
described in section \ref{sec:method}).
We then compared these velocity dispersions, as a function of
\teff, to the velocity dispersions of stars with {\it measured} rotation
periods \citep[\ie\ stars that appear in table 1 of][]{mcquillan2014}.
We show the ratio of total \sigmavb\ to McQuillan-\sigmavb\ as a
function of \teff\ in figure \ref{fig:compare}.}
A larger ratio means the rotating star sample is {\it younger}, on average,
than the overall \Kepler\ population, and a ratio of 1 means that the rotating
stars have the {\it same} age distribution as the overall Kepler sample.
Figure \ref{fig:compare} shows that this ratio is largest for G stars and
approaches unity for K and early M dwarfs.
This indicates that the G stars with detectable rotation periods are, on
average, {\it younger} than the total population of G stars in the Kepler
field.
On the other hand, the late K and early M dwarfs with detectable rotation
periods have a similar age distribution to the overall Kepler population which
suggests that \racomment{most of} the oldest K and M dwarfs are represented in
the \mct\ sample.
This result bolsters the evidence that M dwarf rotation periods are measurable
at older ages than G dwarf rotation periods and that G stars become
magnetically inactive and have fewer active surface regions {\it at a younger
age than M dwarfs}.
